State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-04-29 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
the state estimationsection is clarified

The state estimation technique utilises the Unscented Rauch Tung Striebel Smoother (URTSS)~\cite{Sarkka2010}. The URTSS uses the Unscented Kalman Filter (UKF)~\cite{Julier1997} in the forward iteration to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward path to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. An approximation of the state density is made by propagating the so-called sigma points through the state-space model. The sigma points describe the mean and covariance of the state estimate and are calculated using the unscented transform. The sigma vectors $\mathcal X_i$ are calculated by 
\begin{equation}
	\mathcal X_{0}=\bar x 
\end{equation}
\begin{equation}
	\mathcal X_{i}=\bar x+(\sqrt{( L + \lambda)\mathbf P_x})_i \quad i=1, \dots, L 
\end{equation}
\begin{equation}
	\mathcal X_{i}=\bar x-(\sqrt{( L + \lambda)\mathbf P_x})_{i- L} \quad i= L+1, \dots, 2 L 
\end{equation}
where $L$ is the dimension of the state space and the number of sigma points is $2L+1$. The scaling parameter $\lambda$ is defined as 
\begin{equation}
	\lambda=\alpha^2( L+\kappa)- L 
\end{equation}
where $\alpha$, $\beta$ and $\kappa$ are constants (see ~\cite{Haykin2001} for more details). It should be noted that $\bar x$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$ and $\mathbf{P}_x$ represents the corresponding covariance matrix from the filtering or smoothing respectively. The sigma vectors are propagated through the system equations and weighted to form the posterior mean and covariance. The weights are calculated by 
\begin{equation}
	\mathbf W_0^{(m)}=\frac{\lambda}{ L+\lambda} 
\end{equation}
\begin{equation}
	\mathbf W_0^{(c)}=\frac{\lambda}{ L+\lambda}+(1-\alpha^2+\beta) 
\end{equation}
\begin{equation}
	\mathbf W_i^{(m)}=\mathbf W_i^{(c)}=\frac{1}{2( L+\lambda)} \quad i=1, \dots, 2L. 
\end{equation}

% For explanation of $\lambda, \quad \alpha\quad\text{and}\quad\beta$ see \cite{wan2000unscented,Haykin2001}. $ (\sqrt{(\mathbf L + \lambda)\mathbf P_x})_i$ is the ith column of the matrix square root (e.g., lower-triangular Cholesky factorization). These sigma points propogated through the nonlinear function
The update step equations are the same as the standard Kalman filtering formulations. The state estimates from the forward filtering are used to form a new set of sigma points for the smoother, as described above. However, the predicted cross-covariance of the states needs to be calculated. A summary of the URTSS procedure is given in algorithm~\ref{UKFAlgorithm}. 
\begin{algorithm}
	\caption{The Unscented Kalman Filter}\label{UKFAlgorithm} 
	\begin{algorithmic}[1] 
		\State Forward initialisation 
		\begin{equation*}
			\hat{\mathbf x}_0, \mathbf P_0 
		\end{equation*}
		\State Forward iteration For $t \in \left\lbrace 0,\cdots, T\right\rbrace $, 
		\begin{equation*}
			\mathcal X_t^f=[\hat{\mathbf x}_t^f \quad \hat{\mathbf x}_t^f+\gamma\sqrt{\mathbf P_t^f} \quad \hat{\mathbf x}_t^{f}-\gamma\sqrt{\mathbf P_t^f}] 
		\end{equation*}
		Propagate the sigma points through the state equation 
		\begin{equation}
			\mathcal X_{i,t+1}^{f-}=Q(\mathcal X_t^f,u_t) 
		\end{equation}
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{f-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{f-} 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t +1}^{f-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})^\top+\boldsymbol \Sigma_e 
		\end{equation*}
		\begin{equation*}
			\mathcal K_{t+1}=\mathbf P_{t +1}^{f-}\mathbf C ^\top(\mathbf C \mathbf P_{t +1}^{f-}\mathbf C ^\top+\boldsymbol \Sigma_v)^{-1} 
		\end{equation*}
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{f}=\hat{\mathbf x}_{t+1}^{f-}+\mathcal K_{t+1}(\mathbf y_{t+1}-\mathbf C\hat{\mathbf x}_{t +1}^{f-}) 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t+1}^f=(\mathbf I - \mathcal K_{t+1}\mathbf C)\mathbf P_{t +1}^{f-} 
		\end{equation*}
		\State Backward initialisation 
		\begin{equation}
			\mathbf P_T^b= \mathbf P_T^f 
		\end{equation}
		\begin{equation}
			\hat{\mathbf x}^b_T= \hat{\mathbf x}^f_T 
		\end{equation}
		\State Backward iteration For $t \in \left\lbrace T-1, \cdots, 0 \right\rbrace $ 
		\begin{equation*}
			\mathcal X_t^b=[\hat{\mathbf x}_t^f \quad \hat{\mathbf x}_t^f+\gamma\sqrt{\mathbf P_t^f} \quad \hat{\mathbf x}_t^f-\gamma\sqrt{\mathbf P_t^f}] 
		\end{equation*}
		\State Calculate sigma points for the smoother $\mathcal X_{i,t+1}^{b-}$ and compute 
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{b-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{b-} 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t +1}^{b-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{f})^\top+\boldsymbol \Sigma_e 
		\end{equation*}
		\begin{equation*}
			\mathbf M_{t +1}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t}^{b-}-\hat{\mathbf x}_{t}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t+1}^{f})^\top 
		\end{equation*}
		\State Compute the smoother gain $\mathbf S_t$, the smoothed state $\hat{\mathbf x}_t^b $ and the covariance $\mathbf P_{t}^{b} $ 
		\begin{equation*}
			\mathbf S_t=\mathbf M_{t +1}\left[ \mathbf P_{t +1}^{b-}\right] ^{-1} 
		\end{equation*}
		\begin{equation*}
			\hat{\mathbf x}_t^b=\hat{\mathbf x}_t^f+\mathbf S_t\left[\hat{\mathbf x}_{t+1}^{b}-\hat{\mathbf x}_{t+1}^{b-}\right] 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t}^{b}=\mathbf P_{t}^{f}+\mathbf S_t\left[\mathbf P_{t+1}^{b}-\mathbf P_{t+1}^{b-} \right]\mathbf S_t^\top 
		\end{equation*}
	\end{algorithmic}
\end{algorithm}

>>>>

The state estimation technique utilises the Unscented Rauch Tung Striebel Smoother (URTSS)~\cite{Sarkka2010}. The URTSS uses the Unscented Kalman Filter (UKF)~\cite{Julier1997,Merwe2003} in the forward iteration to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward path to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. An approximation of the state density is made by propagating the so-called sigma points through the state-space model. The sigma points describe the mean and covariance of the state estimate and are calculated using the unscented transform. The sigma vectors $\mathcal X_i$ are calculated by 
\begin{equation}\label{eq:sigmapoints1}
	\mathcal X_{0}=\bar x 
\end{equation}
\begin{equation}
	\mathcal X_{i}=\bar x+(\sqrt{( L + \lambda)\mathbf P_x})_i \quad i=1, \dots, L 
\end{equation}
\begin{equation}\label{eq:sigmapoints2}
	\mathcal X_{i}=\bar x-(\sqrt{( L + \lambda)\mathbf P_x})_{i- L} \quad i= L+1, \dots, 2 L 
\end{equation}
where $(\sqrt{( L + \lambda)\mathbf P_x})_i$ is the ith column of the weighted matrix squre root of  $\mathbf P_x$ and  $L$ is the dimension of the state space. The total number of sigma points is $2L+1$. The scaling parameter $\lambda$ is defined as 
\begin{equation}\label{eq:sigmapoints3}
	\lambda=\alpha^2( L+\kappa)- L 
\end{equation}
where $\alpha$, $\beta$ and $\kappa$ are constants (see ~\cite{Haykin2001} for more details). It should be noted that $\bar x$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$ and $\mathbf{P}_x$ represents the corresponding covariance matrix from the filtering or smoothing respectively. The sigma vectors are propagated through the system equations and weighted to form the posterior mean and covariance. The weights are calculated by 
\begin{equation}
	\mathbf W_0^{(m)}=\frac{\lambda}{ L+\lambda} 
\end{equation}
\begin{equation}
	\mathbf W_0^{(c)}=\frac{\lambda}{ L+\lambda}+(1-\alpha^2+\beta) 
\end{equation}
\begin{equation}
	\mathbf W_i^{(m)}=\mathbf W_i^{(c)}=\frac{1}{2( L+\lambda)} \quad i=1, \dots, 2L. 
\end{equation}

% For explanation of $\lambda, \quad \alpha\quad\text{and}\quad\beta$ see \cite{wan2000unscented,Haykin2001}. $ (\sqrt{(\mathbf L + \lambda)\mathbf P_x})_i$ is the ith column of the matrix square root (e.g., lower-triangular Cholesky factorization). These sigma points propogated through the nonlinear function
Since the observation equation \ref{ObservationEquation} is linear the update step equations are simplified to the standard Kalman filtering formulations. The state estimates from the forward filtering are used to form a new set of sigma points for the smoother, as described above. However, the predicted cross-covariance of the states needs to be calculated. A summary of the URTSS procedure is given in algorithm~\ref{UKFAlgorithm}. It should be noted that as the disturbance, $e_t\left(\mathbf{r}\right)$ and the measurement noise $ \boldsymbol{\varepsilon}_t$ are additive the additive form of the URTS can be used. This reduces the dimension and the number of sigma points used in the algorithm.
\begin{algorithm}
	\caption{The Unscented Kalman Filter}\label{UKFAlgorithm} 
	\begin{algorithmic}[1] 
		\State Forward initialisation 
		\begin{equation*}
			\hat{\mathbf x}_0, \mathbf P_0 
		\end{equation*}
		\State Forward iteration For $t \in \left\lbrace 0,\cdots, T\right\rbrace $,

		calculate the sigma points $\mathcal X_{i,t}^f$ using \ref{eq:sigmapoints1}-\ref{eq:sigmapoints3} and propogate them through the state equation
% 		\begin{equation*}
% 			\mathcal X_t^f=[\hat{\mathbf x}_t^f \quad \hat{\mathbf x}_t^f+\gamma\sqrt{\mathbf P_t^f} \quad \hat{\mathbf x}_t^{f}-\gamma\sqrt{\mathbf P_t^f}] 
% 		\end{equation*}

% 		propagate the sigma points through the state equation 
		\begin{equation}
			\mathcal X_{i,t+1}^{f-}=Q(\mathcal X_{i,t}^f,u_t) 
		\end{equation}

		calculate the predicted mean and the predicted covariance matrix
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{f-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{f-} 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t +1}^{f-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})^\top+\boldsymbol \Sigma_e 
		\end{equation*}

		compute the filter gain, the update mean and the update covariance matrix using the standard kalman filter formulations
		\begin{equation*}
			\mathcal K_{t+1}=\mathbf P_{t +1}^{f-}\mathbf C ^\top(\mathbf C \mathbf P_{t +1}^{f-}\mathbf C ^\top+\boldsymbol \Sigma_v)^{-1} 
		\end{equation*}
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{f}=\hat{\mathbf x}_{t+1}^{f-}+\mathcal K_{t+1}(\mathbf y_{t+1}-\mathbf C\hat{\mathbf x}_{t +1}^{f-}) 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t+1}^f=(\mathbf I - \mathcal K_{t+1}\mathbf C)\mathbf P_{t +1}^{f-} 
		\end{equation*}
		\State Backward initialisation 
		\begin{equation}
			\mathbf P_T^b= \mathbf P_T^f 
		\end{equation}
		\begin{equation}
			\hat{\mathbf x}^b_T= \hat{\mathbf x}^f_T 
		\end{equation}
		\State Backward iteration For $t \in \left\lbrace T-1, \cdots, 0 \right\rbrace $

calculate the sigma points $\mathcal X_{i,t}^b$ using \ref{eq:sigmapoints1}-\ref{eq:sigmapoints3} and propogate them through the state equation
		\begin{equation}
			\mathcal X_{i,t+1}^{b-}=Q(\mathcal X_{i,t}^b,u_t) 
		\end{equation}
% 		\begin{equation*}
% 			\mathcal X_t^b=[\hat{\mathbf x}_t^f \quad \hat{\mathbf x}_t^f+\gamma\sqrt{\mathbf P_t^f} \quad \hat{\mathbf x}_t^f-\gamma\sqrt{\mathbf P_t^f}] 
% 		\end{equation*}
		\State compute the predicted mean, the predicted covariance matrix and the cross-covariance matrix
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{b-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{b-} 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t +1}^{b-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{f})^\top+\boldsymbol \Sigma_e 
		\end{equation*}
		\begin{equation*}
			\mathbf M_{t +1}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t}^{b-}-\hat{\mathbf x}_{t}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t+1}^{f})^\top 
		\end{equation*}
		\State Compute the smoother gain $\mathbf S_t$, the smoothed state $\hat{\mathbf x}_t^b $ and the covariance $\mathbf P_{t}^{b} $ 
		\begin{equation*}
			\mathbf S_t=\mathbf M_{t +1}\left[ \mathbf P_{t +1}^{b-}\right] ^{-1} 
		\end{equation*}
		\begin{equation*}
			\hat{\mathbf x}_t^b=\hat{\mathbf x}_t^f+\mathbf S_t\left[\hat{\mathbf x}_{t+1}^{b}-\hat{\mathbf x}_{t+1}^{b-}\right] 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t}^{b}=\mathbf P_{t}^{f}+\mathbf S_t\left[\mathbf P_{t+1}^{b}-\mathbf P_{t+1}^{b-} \right]\mathbf S_t^\top 
		\end{equation*}
	\end{algorithmic}
\end{algorithm}
